Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2022-01-08 analysis/paper/paper.Rmd
made a few writing edits

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our initial intent was to sample during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span, but due to budget constraints the spring 2021 effort was limited to a single week. Fall efforts were constrained within a two-week window to minimize predator emigration/immigration and to constrain the population for the mark-recapture models.

@@

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer")) %>%
  arrange(event_name, model)

@@

where $M$, $n$, and $m$ are indexed by each sample, $i$. Here, sample $i$ is defined as each day. The Schnabel estimator does not have an associated standard error; however, the 95% confidence intervals were calculated. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Champman modification). Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Several assumptions in our estimators may have influenced the magnitude of our results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, we did not observe mortalities or injury post release; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that the second (and subsequent) sample(s) are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size threshold susceptible to angling. We also consider the validity of our assumption of equal catchability of individual fish between sampling events to have introduced potential bias in our results. Unequal catchability for individual fish is difficult to assess, especially if being caught once influences the likelihood of recapture. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> We further recognize that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. If marked fish were not recently released or mixed back into those areas that would result in marked fish being less susceptible to capture. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability.

>>>>

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. Our initial intent was to sample during the emigration of the 2018 brood year (fall 2019 and spring 2020). However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 effort was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional effort was added in spring 2021. Fall efforts included sampling across a two-week span, but due to budget constraints the spring 2021 effort was limited to a single week. Fall efforts were constrained within a two-week window to minimize predator emigration/immigration and to constrain the population to meet assumptions of the mark-recapture models.

@@

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer"),
         model = fct_relevel(model,
                             "Petersen",
                             after = 0)) %>%
  arrange(event_name, model)

@@

where $M$, $n$, and $m$ are indexed by each sample (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999]. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Peterson estimators (with a Champman modification). Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

@@

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Several assumptions in our estimators may have influenced the magnitude of our results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, we did not observe mortalities or injury post release; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that the second (and subsequent) sample(s) are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size threshold susceptible to angling, and possibly biasing them low. We also consider the validity of our assumption of equal catchability of individual fish between sampling events to have introduced potential bias in our results. Unequal catchability for individual fish is difficult to assess, especially if being caught once influences the likelihood of recapture. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> We further recognize that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. If marked fish were not recently released or mixed back into those areas that would result in marked fish being less susceptible to capture. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability.
